---
title: "The COVID-19 pandemic's gendered impact on academic productivity"
author: "Megan Frederickson"
date: "18/04/2020"
output: github_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# How is COVID-19 affecting the productivity of male versus female academics?

There is debate on Twitter about how the COVID-19 pandemic is affecting submissions to journals, and expecially whether there are differences between male and female scholars. There are anectodal observations that submissions from female scholars are down in March, 2020, presumably because women are disproportionately shouldering increased caregiving burdens because of school and childcare closures. I wanted to find out it these anecdotes are borne out by data. 

I don't not have access to journal submission data directly, so I decided to use data from large preprint servers, where academics often archive their papers as they submit them for peer-review to journals.

First, I loaded some useful packages. 

```{r Load packages, message=FALSE, warning=FALSE}

library(tidyverse) #includes ggplot2, dplyr, readr, stringr
library(knitr)
library(cowplot)
library(car)
library(nlme)
library(lme4)
library(glmm)
library(gender)
library(png)
library(reticulate)
library(aRxiv)
library(lubridate)
library(anytime)
library(zoo)
library(viridis)

```

Next, I scraped submission data from arXiv (https://arxiv.org/), which is the main preprint server for physics, math, computer science, statistics, and other quantitative disciplines. I scraped all records for March 15-April 15, 2020, during the COVID-19 pandemic, and for the same date range in 2019.

```{r Scrape arxiv, eval=FALSE}
#Not run
#Get all submissions between March 15, 2020 and April 15, 2020 (during the COVID-19 pandemic)
n.2020 <- arxiv_count(query = 'submittedDate:[20200315 TO 20200415]')
df.2020.1 <- arxiv_search(query = 'submittedDate:[20200315 TO 20200321]', limit=n.2020, batchsize=1000)
df.2020.2 <- arxiv_search(query = 'submittedDate:[20200322 TO 20200328]', limit=n.2020, batchsize=1000)
df.2020.3 <- arxiv_search(query = 'submittedDate:[20200329 TO 20200403]', limit=n.2020, batchsize=1000)
df.2020.4 <- arxiv_search(query = 'submittedDate:[20200404 TO 20200409]', limit=n.2020, batchsize=1000)
df.2020.5 <- arxiv_search(query = 'submittedDate:[20200410 TO 20200412]', limit=2000, batchsize=500)
df.2020.6 <- arxiv_search(query = 'submittedDate:[20200413 TO 20200415]', limit=2000, batchsize=500)
df.2020.full <- rbind(df.2020.1, df.2020.2, df.2020.3, df.2020.4, df.2020.5, df.2020.6)
write.csv(df.2020.full, file="arxiv_2020_data.csv")

#Get all submission between March 15, 2019 and April 15, 2019 (the same dates last year)
n.2019 <- arxiv_count(query = 'submittedDate:[20190315 TO 20190415]')
df.2019.1 <- arxiv_search(query = 'submittedDate:[20190315 TO 20190322]', limit=n.2019, batchsize=1000)
df.2019.2 <- arxiv_search(query = 'submittedDate:[20190323 TO 20190329]', limit=n.2019, batchsize=1000)
df.2019.3 <- arxiv_search(query = 'submittedDate:[20190330 TO 20190405]', limit=n.2019, batchsize=1000)
df.2019.4 <- arxiv_search(query = 'submittedDate:[20190406 TO 20190412]', limit=n.2019, batchsize=1000)
df.2019.5 <- arxiv_search(query = 'submittedDate:[20190413 TO 20190415]', limit=n.2019, batchsize=500)
df.2019.full <- rbind(df.2019.1, df.2019.2, df.2019.3, df.2019.4, df.2019.5)
write.csv(df.2019.full, file="arxiv_2019_data.csv")
```

I then tidied the data by splitting strings of author names to identify the first and last author of each preprint. I also calculated the number of authors per preprint. 

```{r Tidy author names}

df.2020 <- read.csv("arxiv_2020_data.csv") #Read in data
df.2019 <- read.csv("arxiv_2019_data.csv")

df.2020$author.n <- str_count(df.2020$authors, pattern = "\\|")+1 #Count author number
df.2019$author.n <- str_count(df.2019$authors, pattern = "\\|")+1 

max(df.2020$author.n) #Max author number 
max(df.2019$author.n) 

kable(df.2020 %>% group_by(author.n) %>% summarize(n=n()))
kable(df.2019 %>% group_by(author.n) %>% summarize(n=n()))

df.2020$first.author <- sapply(strsplit(as.character(df.2020$authors), "|", fixed=TRUE), head, 1) #Extract first author 
df.2019$first.author <- sapply(strsplit(as.character(df.2019$authors), "|", fixed=TRUE), head, 1)

df.2020$last.author <- sapply(strsplit(as.character(df.2020$authors), "|", fixed=TRUE), tail, 1) #Extract last author
df.2019$last.author <- sapply(strsplit(as.character(df.2019$authors), "|", fixed=TRUE), tail, 1) 

df.2020$first.author.first.name <- sapply(strsplit(df.2020$first.author, " "), head, 1) #Extract first author first name
df.2019$first.author.first.name <- sapply(strsplit(df.2019$first.author, " "), head, 1)

df.2020$last.author.first.name <- sapply(strsplit(df.2020$last.author, " "), head, 1) #Extract last author first name
df.2019$last.author.first.name <- sapply(strsplit(df.2019$last.author, " "), head, 1)

#all.authors.2020 <- unlist(strsplit(as.character(df.2020$authors), "|", fixed=TRUE))
#all.authors.first.names.2020 <- sapply(strsplit(all.authors.2020, " "), head, 1)

#all.authors.2019 <- unlist(strsplit(as.character(df.2019$authors), "|", fixed=TRUE))
#all.authors.first.names.2019 <- sapply(strsplit(all.authors.2019, " "), head, 1)

```

Then I cleaned up the categories and dates for later use. 

```{r Tidy dates and categories}

df.2020$cat <- sapply(strsplit(as.character(df.2020$primary_category), ".", fixed=TRUE), head, 1) #Lump similar categories
df.2019$cat <- sapply(strsplit(as.character(df.2019$primary_category), ".", fixed=TRUE), head, 1)

df.2019$date <- as.Date(df.2019$submitted)
df.2020$date <- as.Date(df.2020$submitted)

```

I assigned gender to author first names using the gender package (https://github.com/ropensci/gender). This package returns the probability that a name is male or female by comparing the name to a names in a database; I used the U.S. Social Security baby names database. I started by assigning gender to first and last authors of preprints.

Please note: this is a brute force and rather crude method of predicting gender, and it has many limitations, as discussed by the package authors on their GitHub repo and included links. By using this method, I am not assuming that individuals are correctly gendered in the resulting dataset, but merely that it provides insight into gender's effects across the aggregated population of preprint authors. 

```{r Assign first and last author gender}

gender.2020 <- NULL
gender.2020 <- gender(df.2020$first.author.first.name, method = "ssa")
gender.2020 <- unique(gender.2020[ , c(1,2,4)])
df.2020 <- merge(df.2020, gender.2020, by.x = "first.author.first.name",  by.y ="name", all = TRUE)
colnames(df.2020)[which(colnames(df.2020) == "proportion_male")] <- "first.author.prop.male"
colnames(df.2020)[which(colnames(df.2020) == "gender")] <- "first.author.gender"

gender.2020 <- NULL
gender.2020 <- gender(df.2020$last.author.first.name, method = "ssa")
gender.2020 <- unique(gender.2020[ , c(1,2,4)])
df.2020 <- merge(df.2020, gender.2020, by.x = "last.author.first.name",  by.y ="name", all = TRUE)
colnames(df.2020)[which(colnames(df.2020) == "proportion_male")] <- "last.author.prop.male"
colnames(df.2020)[which(colnames(df.2020) == "gender")] <- "last.author.gender"

gender.2019 <- NULL
gender.2019 <- gender(df.2019$first.author.first.name, method = "ssa")
gender.2019 <- unique(gender.2019[ , c(1,2,4)])
df.2019 <- merge(df.2019, gender.2019, by.x = "first.author.first.name",  by.y ="name", all = TRUE)
colnames(df.2019)[which(colnames(df.2019) == "proportion_male")] <- "first.author.prop.male"
colnames(df.2019)[which(colnames(df.2019) == "gender")] <- "first.author.gender"

gender.2019 <- NULL
gender.2019 <- gender(df.2019$last.author.first.name, method = "ssa")
gender.2019 <- unique(gender.2019[ , c(1,2,4)])
df.2019 <- merge(df.2019, gender.2019, by.x = "last.author.first.name",  by.y ="name", all = TRUE)
colnames(df.2019)[which(colnames(df.2019) == "proportion_male")] <- "last.author.prop.male"
colnames(df.2019)[which(colnames(df.2019) == "gender")] <- "last.author.gender"

```

Next, I predicted the genders of ALL preprint authors, and summarized the data as the number of male and female authors of each preprint, regardless of order. This code takes a while to run, so it is not run when knitting this markdown document.

```{r, Assign all authors gender, eval=FALSE}

#Not run
split.names <- function(x){strsplit(as.character(x), "|", fixed=TRUE)}  
df.2020$split.names <- lapply(df.2020$authors, split.names)
tmp <- NULL
for(i in 1:length(df.2020$authors)){
  tmp[i] <- as.vector(gender(word(unlist(df.2020$split.names[[i]]), 1), method="ssa")[, 4])
  df.2020$male.n[i] <- as.numeric(str_count(as.character(tmp[i]), pattern = paste(sprintf("\\b%s\\b", "male"))))
  df.2020$female.n[i] <- as.numeric(str_count(as.character(tmp[i]), pattern = paste(sprintf("\\b%s\\b", "female"))))
}
df.2020$prop.female.authors <- df.2020$female.n/(df.2020$female.n+df.2020$male.n)
df.2020$prop.female.authors.ex.single <- ifelse((df.2020$female.n+df.2020$male.n) > 1, df.2020$prop.female.authors, NA)
df.2020.output <- as.data.frame(apply(df.2020,2,as.character))
write.csv(df.2020.output, "arxiv_2020_gender.csv")

split.names <- function(x){strsplit(as.character(x), "|", fixed=TRUE)}  
df.2019$split.names <- lapply(df.2019$authors, split.names)
tmp <- NULL
for(i in 1:
length(df.2019$authors)){
  tmp[i] <- as.vector(gender(word(unlist(df.2019$split.names[[i]]), 1), method="ssa")[, 4])
  df.2019$male.n[i] <- as.numeric(str_count(as.character(tmp[i]), pattern = paste(sprintf("\\b%s\\b", "male"))))
  df.2019$female.n[i] <- as.numeric(str_count(as.character(tmp[i]), pattern = paste(sprintf("\\b%s\\b", "female"))))
}

df.2019$prop.female.authors <- df.2019$female.n/(df.2019$female.n+df.2019$male.n)
df.2019$prop.female.authors.ex.single <- ifelse((df.2019$female.n+df.2019$male.n) > 1, df.2019$prop.female.authors, NA)
df.2019.output <- as.data.frame(apply(df.2019,2,as.character))
write.csv(df.2019.output, "arxiv_2019_gender.csv")

```

## First authors

I calculated what proportion of first authors are female, and then visualized this proportion over time, comparing the same date range in 2019 and 2020. There appears to be no differences, at least aggregated across all arxiv preprints (e.g., without considering field, etc.)

```{r Visualize first author data, fig.width=4, fig.height=4, dpi=600}

df.2020 <- read.csv("arxiv_2020_gender.csv") #Read in data
df.2019 <- read.csv("arxiv_2019_gender.csv")

#Rerun
df.2019$date <- as.Date(df.2019$submitted)
df.2020$date <- as.Date(df.2020$submitted)

sum.2019 <- subset(df.2019, !is.na(first.author.gender)) %>% group_by(date) %>% summarize(n=n(), female.n = length(which(first.author.gender =="female")), female.per = female.n/n)
sum.2019 <- ungroup(sum.2019)
sum.2019$year <- "2019"
sum.2019$date.noyear <- paste0(month(sum.2019$date), "/", day(sum.2019$date))

sum.2020 <- subset(df.2020, !is.na(first.author.gender)) %>% group_by(date) %>% summarize(n=n(), female.n = length(which(first.author.gender =="female")), female.per = female.n/n)
sum.2020 <- ungroup(sum.2020)
sum.2020$year <- "2020"
sum.2020$date.noyear <- paste0(month(sum.2020$date), "/", day(sum.2020$date))

summary <- rbind(sum.2019, sum.2020)
summary$anydate <- anydate(summary$date.noyear)

p1 <- ggplot(data=summary, aes(x=as.Date(anydate), y=female.per, color=year))+geom_point()+ylab("Female (%)")+xlab("Date")+labs(color="Year")+geom_smooth(method="lm", se=FALSE)+ggtitle("First authors")+theme_cowplot()
p1

```

## Last authors

I then visualized what proportion of last authors are female and how this changes over time, again comparing the same date range in 2019 and 2020. Here, there does appear to be fewer female last authors in 2020 than in 2019. 

```{r Visualize last author data, fig.width=4, fig.height=4, dpi=600}

sum.2019 <- subset(df.2019, !is.na(last.author.gender)) %>% group_by(date) %>% summarize(n=n(), female.n = length(which(last.author.gender =="female")), female.per = female.n/n)
sum.2019 <- ungroup(sum.2019)
sum.2019$year <- "2019"
sum.2019$date.noyear <- paste0(month(sum.2019$date), "/", day(sum.2019$date))

sum.2020 <- subset(df.2020, !is.na(last.author.gender)) %>% group_by(date) %>% summarize(n=n(), female.n = length(which(last.author.gender =="female")), female.per = female.n/n)
sum.2020 <- ungroup(sum.2020)
sum.2020$year <- "2020"
sum.2020$date.noyear <- paste0(month(sum.2020$date), "/", day(sum.2020$date))

summary <- rbind(sum.2019, sum.2020)
summary$anydate <- anydate(summary$date.noyear)
summary$weekday <- weekdays(summary$date)

p2 <- ggplot(data=summary, aes(x=anydate, y=female.per, color=year))+geom_point()+ylab("Female (%)")+xlab("Date")+labs(color="Year")+geom_smooth(method="lm", se=FALSE)+ggtitle("Last authors")+theme_cowplot()
p2

p3 <- plot_grid(p1, p2, nrow=2, labels=c("A", "B"))
p3
save_plot("plot.png", p3, base_height=5, base_width=6)

```

I was curious how this effect plays out by field, so I aggregated the category information provided by arXiv into six main categories: physics, math, stats, computer science, economics, and electrical engineering. Only three categories have enough records to really use reliably: physics, math, and computer science. It looks like the decline in female last authorship in 2020 is driven in large part by physics and math, but not computer science. 

```{r Last authors by field, fig.width=4, fig.height=4, dpi=600}

df.full <- rbind(df.2019, df.2020)
df.full$year <- as.factor(year(df.full$date))
df.full$big.cat <- ifelse((df.full$cat == "astro-ph" | df.full$cat == "cond-mat" | df.full$cat == "gr-qc" | df.full$cat == "hep-ex" | df.full$cat == "physics" | df.full$cat == "nucl-ex"| df.full$cat == "quant-ph" | df.full$cat == "nucl-th" | df.full$cat == "hep-ph" | df.full$cat == "nlin" | df.full$cat == "hep-th" | df.full$cat == "hep-lat"), "physics", ifelse(df.full$cat == "stat", "stats", ifelse(df.full$cat == "math" | df.full$cat == "math-ph", "math", ifelse(df.full$cat=="cs", "cs", ifelse(df.full$cat == "q-bio", "q-bio", ifelse(df.full$cat=="q-fin" | df.full$cat == "econ", "econ", ifelse(df.full$cat=="eess", "eess", NA)))))))

big.sum.last <- subset(df.full, !is.na(last.author.gender)) %>% group_by(year, big.cat) %>% summarize(total = n(), female.n = length(which(last.author.gender =="female")), female.per = female.n/total)

big.sum.first <- subset(df.full, !is.na(first.author.gender)) %>% group_by(year, big.cat) %>% summarize(total = n(), female.n = length(which(first.author.gender =="female")), female.per = female.n/total)

min = 500

p4 <- ggplot()+geom_point(data=subset(big.sum.last, total > min), aes(x=year,y=female.per), size=5, color="blue")+geom_line(data=subset(big.sum.last, total > min), aes(x=year,y=female.per, group=1), size=1, color="blue")+facet_wrap(~big.cat)
p4

```

## All authors

Next I wanted to look at authorship in general, because we see different patterns for first and last authors, but conventions regarding author order are not consistent across fields.

### Sole authors

How many preprints were sole-authored by a male versus a female academic in Mar/Apr, 2020, compared to the same dates last year? 

```{r Sole authors}

summary <- df.full %>% group_by(year, male.n, female.n) %>% summarize(n = n())
summary$total.n <- summary$male.n+summary$female.n
summary.sole.authors <- subset(summary, total.n == 1)
sole.wide <- spread(summary.sole.authors, year, n)
sole.wide$gender <- ifelse(sole.wide$male.n == 0, "Female", "Male")
sole.long <- gather(sole.wide, year, number, `2019`:`2020`)
sole.wide$per.dif <- (sole.wide$`2020`/sole.wide$`2019`)*100-100

p5 <- ggplot(data=sole.long, aes(fill=as.factor(year), y=number, x=gender))+geom_bar(position="dodge", stat="identity")+theme_cowplot()+xlab("Gender")+ylab("Preprints (no.)")+labs(fill="Year")+scale_fill_viridis(option="cividis", discrete=TRUE)+annotate("text", x=c(1,2),  y=c(650,2600), label = paste0("+", round(sole.wide$per.dif, 1), "%"))+ggtitle("1 author")+theme(legend.position = c(0.1, 0.85))
p5

```

### Two authors

```{r All authors}

summary.two.authors <- subset(summary, total.n == 2)
two.wide <- spread(summary.two.authors, year, n)
two.wide$gender <- ifelse(two.wide$male.n == 0, "All female", ifelse(two.wide$female.n == 0, "All male", "Mixed"))
two.long <- gather(two.wide, year, number, `2019`:`2020`)
two.wide$per.dif <- (two.wide$`2020`/two.wide$`2019`)*100-100

p6 <- ggplot(data=two.long, aes(fill=as.factor(year), y=number, x=gender))+geom_bar(position="dodge", stat="identity")+theme_cowplot()+xlab("Gender")+ylab("Preprints (no.)")+labs(fill="Year")+scale_fill_viridis(option="cividis", discrete=TRUE)+ggtitle("2 authors")+theme(legend.position = "none")
p6

```

### Three or more authors

```{r}

summary.three.authors <- subset(summary, total.n >= 3)
three.wide <- spread(summary.three.authors, year, n)
three.wide$gender <- ifelse(three.wide$male.n == 0, "All female", ifelse(three.wide$female.n == 0, "All male", "Mixed"))
three.wide.sum <- three.wide %>% group_by(gender) %>% summarize(sum.2019 = sum(`2019`, na.rm=TRUE), sum.2020 = sum(`2020`, na.rm=TRUE))
three.long <- gather(three.wide.sum, year, number, sum.2019:sum.2020)

p7 <- ggplot(data=three.long, aes(fill=as.factor(year), y=number, x=gender))+geom_bar(position="dodge", stat="identity")+theme_cowplot()+xlab("Gender")+ylab("Preprints (no.)")+labs(fill="Year")+scale_fill_viridis(option="cividis", discrete=TRUE)+ggtitle("3+ authors")+theme(legend.position = "none")
p7

p8 <- plot_grid(p5,p6,p7, labels=c("A", "B", "C"))
save_plot("plot2.png", p8, base_height=6, base_width=8)
```

```{r}
summary$female.n <- as.factor(summary$female.n)
sum.sum.female <- subset(summary, total.n > 0 & female.n != 0) %>% group_by (year, female.n) %>% summarize(total.pubs = sum(n, na.rm=TRUE))
sum.sum.female.wide <- spread(sum.sum.female, year, total.pubs)
sum.sum.female.wide$female.n <- as.numeric(as.character(sum.sum.female.wide$female.n))
sum.sum.female.wide$female.n.trunc <- ifelse(sum.sum.female.wide$female.n >= 5, "5+", sum.sum.female.wide$female.n)
sum.sum.female.wide$female.n.trunc <- as.factor(sum.sum.female.wide$female.n.trunc)
female <- sum.sum.female.wide %>% group_by(female.n.trunc) %>% summarize(sum.2019 = sum(`2019`, na.rm=TRUE), sum.2020 = sum (`2020`, na.rm=TRUE))
levels(female$female.n.trunc) <- c("1", "2", "3", "4", "5+", "Overall")
female <- rbind(female, c("Overall", sum(female$sum.2019), sum(female$sum.2020)))
female$per.dif <- (as.numeric(female$sum.2020)/as.numeric(female$sum.2019))*100-100
female$gender <-"Female"
colnames(female)[1] <- "Male or female authors (no.)"

summary$male.n <- as.factor(summary$male.n)
sum.sum.male <- subset(summary, total.n > 0 & male.n != 0) %>% group_by (year, male.n) %>% summarize(total.pubs = sum(n, na.rm=TRUE))
sum.sum.male.wide <- spread(sum.sum.male, year, total.pubs)
sum.sum.male.wide$male.n <- as.numeric(as.character(sum.sum.male.wide$male.n))
sum.sum.male.wide$male.n.trunc <- ifelse(sum.sum.male.wide$male.n >= 5, "5+", sum.sum.male.wide$male.n)
sum.sum.male.wide$male.n.trunc <- as.factor(sum.sum.male.wide$male.n.trunc)
male <- sum.sum.male.wide %>% group_by(male.n.trunc) %>% summarize(sum.2019 = sum(`2019`, na.rm=TRUE), sum.2020 = sum (`2020`, na.rm=TRUE))
levels(male$male.n.trunc) <- c("1", "2", "3", "4", "5+", "Overall")
male <- rbind(male, c("Overall", sum(male$sum.2019), sum(male$sum.2020)))
male$per.dif <- (as.numeric(male$sum.2020)/as.numeric(male$sum.2019))*100-100
male$gender <- "Male"
colnames(male)[1] <- "Male or female authors (no.)"

all <- rbind(female, male)
all$dif <- as.numeric(all$sum.2020) - as.numeric(all$sum.2019)
all.long <- gather(all, year, number, sum.2019:sum.2020)
all.long$year <- as.factor(all.long$year)
levels(all.long$year) <- c("2019", "2020")


p9 <- ggplot(data=all, aes(fill=gender, x=`Male or female authors (no.)`, y=per.dif))+ylab("Percent change, 2019 to 2020")+geom_bar(position="dodge", stat="identity")+theme_cowplot()
p9

p10 <- ggplot(data=all, aes(fill=gender, x=`Male or female authors (no.)`, y=dif))+ylab("Change in number of preprints, 2019 to 2020")+geom_bar(position="dodge", stat="identity")+theme_cowplot()
p10

p11 <- ggplot(data=all.long, aes(fill=year, x=`Male or female authors (no.)`, y=as.numeric(number)))+labs(fill="Year")+ylab("Preprints (no.)")+geom_bar(position="dodge", stat="identity")+theme_cowplot()+scale_fill_viridis(discrete=TRUE)+facet_wrap(~gender)
p11
```

